---
title: "TQIP's Risk-Adjusted Boxplots"
author: "Jake Roden-Foreman"
date: '2021-08-11'
output:
  md_document:
    variant: gfm
    preserve_yaml: yes
excerpt: The boxplots in TQIP's risk-adjusted benchmark reports are confusing.  I try to explain them.
tags:
- Trauma
- TQIP
- R
- Data Visualization
editor_options: 
  chunk_output_type: console
knit: (function(inputFile, encoding) { rmarkdown::render(inputFile, encoding = encoding,
  output_dir = "../_posts") })
toc: true
---


```{r setup, include=FALSE}


base_dir <- "~/Desktop/traumadata.github.io/"
base_url <- "/"


#save plots in the folder for this post in the "img/posts/" directory
file_path = rstudioapi::getActiveDocumentContext()$path #the file path of the current file
img_prefix = gsub(x = basename(file_path), 
									pattern = "(\\.Rmd)$", 
									replacement = "")
fig_path <- paste0("img/posts/", img_prefix, "/")




knitr::opts_knit$set(base.dir = base_dir, base.url = base_url)

knitr::opts_chunk$set(fig.path = fig_path, dpi = 150,
											fig.align = "center",
											out.width = '100%',
                      message = FALSE, warning = FALSE,
                      cache = FALSE, echo = FALSE,
											results = "as.is"
											) 

library(tidyverse)
library(ggbeeswarm)
library(gginnards) #to edit/reorder geom layers

theme_set(theme_classic())


```

# Intro

U.S. trauma centers are required to "use a risk-adjusted benchmarking system to measure
performance and outcomes."[^1]   This is most commonly done by submitting registry data to the Trauma Quality Improvement Program (TQIP).  TQIP used to provide caterpillar plots to help visualize how each center performed compared to other centers.  However, caterpillar plots aren't particularly helpful for showing how one specific center did compared to other centers (the whole point of benchmarking); they are better suited for showing the overall pattern/distribution of performance.  

Therefore, TQIP moved to using "modified boxplots" to better highlight the performance of the center in the benchmark report.  I really like these boxplots. They pack a lot of information into a single figure, and that's what good data visualization is all about.  However, they are also *really* confusing for a lot of people because they are so information dense.  

To TQIP's credit, they (and many others) have created numerous publications and presentations trying to explain their boxplots, but there is still a lot of confusion out their.  So here is the inaugural post on the Trauma Data Blog: my attempt to explain TQIP's risk-adjusted boxplots. 


[^1]: [Resources for Optimal Care of the Injured Patient 2014, page 109](https://www.facs.org/-/media/files/quality-programs/trauma/vrc-resources/resources-for-optimal-care.ashx)


# Odds ratios

```{r make-data}
# #find a good seed
# for(i in 1:50){
# 	set.seed(i)
# 	x = rnorm(n=500, mean = 1.05, sd = 0.3)
# 	cat("i=", i, "\tmed=", median(x), " min=", min(x), "\n")
# }



#make data
set.seed(5)
x = rnorm(n=500, mean = 1.05, sd = 0.3)
# summary(x)
# quantile(x, probs = seq(0, 1, 0.1))

mdata = tibble(
	x = x,
	cuts = cut(x, 
				breaks = quantile(x, probs=seq(0, 1, by=0.1), na.rm=TRUE), 
				include.lowest=TRUE),
	decile = cuts %>% as.factor() %>% as.numeric()
	)


mdata = mdata %>% 
	mutate(
	color = case_when(
		decile == 1 ~ alpha("green", 0.5),
		decile == 10 ~ alpha("red", 0.5),
	TRUE ~ alpha("gray", 0.5)
	)
)
```



When TQIP does its benchmark reports, it calculates `risk-adjusted odds ratios` as its primary method of conveying your center's performance.  You can think of these odds ratios as the ratio of the number of events that were <u><em>o</em></u>bserved in the data (that is, the actual number of deaths) relative to the number of events that would be <u><em>e</em></u>xpected or predicted to occur at an average TQIP hospital. (The technical interpretation is a bit more complicated than that, but that description is good enough for this post.) That's why these odds ratios are sometimes called <u><em>O</em></u> to <u><em>E</em></u> ratios.  

Odds ratios greater than one (>1) are unfavorable because they indicate there were *more events at your center* than TQIP's models predicted based on the data.  Similarly, odds ratios less than one (<1) are good because they indicate there were *fewer events at your center* than predicted.  Finally, an odds ratio of one means there were exactly the same number of observed events as there were predicted events.

So, an odds ratio of 1.5 would mean that the odds of the event (whatever it is) were 50% higher (1.5 - 1.0 = 0.5 <span>&#8594;</span> 50%) at your center than expected based on the data.  But, an odds ratio of 0.75 would mean that the odds of the event were 25% lower (1.0 - 0.75 = 0.25 <span>&#8594;</span> 25%) than expected at your center given the data.


\


# Setting the scene

After TQIP calculates each hospital's odds ratio for a given outcome, the distribution of all those odds ratios will (usually) look like a bell curve or a normal curve.  Using some simulated data, the bell curve of all those odds ratios might look like this.  This is a histogram (a type of bar chart) of the simulated odds ratios with a blue curve to smooth over the bumps in the histogram. As you can see, the odds ratios are pretty normally distributed with most of them falling around 1. 

```{r histo-with-dens}
#histogram with density


library(grid)


p = ggplot(mapping = aes(x=x, y = ..density..)) +
	geom_histogram(
		binwidth = 0.1,
		fill = alpha("gray", 0.5),
		color = alpha("gray", 0.5)
		) +
	geom_density(color = "#6EE2FF") +
	labs(x = "Odds Ratio") +
	scale_y_continuous(breaks = NULL, labels = NULL, name = NULL) +
	coord_cartesian(xlim = c(0, max(x))) +
	theme(
		axis.line.y = element_line(arrow = grid::arrow(length = unit(0.3, "cm"), ends = "both")),
		axis.title.y = element_text(angle = 0),
		plot.margin = unit(c(1,1,1,3), "cm")
	) +
	annotate("text", label = "More Common", 
					 x = 0, y = 1.4, hjust = 1.3) +
	annotate("text", label = "Less Common", 
					 x = 0, y = 0, hjust = 1.3) +
	annotate("text", label = "\u2190 Better Performance", 
					 x = 0.25, y = 0, vjust = 6) +
	annotate("text", label = "Worse Performance \u2192", 
					 x = 1.6, y = 0, vjust = 6)



# Code to override clipping
p = ggplot_gtable(ggplot_build(p))
p$layout$clip[p$layout$name == "panel"] <- "off"
grid.draw(p)





```


\



Alternatively, this histogram can be shown by plotting the individual odds ratios and stacking them. Showing the individual odds ratios will help us understand the TQIP boxplots later on. 

```{r histodot, fig.height=5.5}
#histogram made of dots
ggplot(mapping = aes(x=x)) +
	geom_dotplot(
		fill = alpha("gray", 0.5),
		color = alpha("gray", 0.5),
		binwidth = 0.1,
		dotsize = 0.18,
		method = "histodot"
	) +
	labs(x = "Odds Ratio") +
	scale_y_continuous(breaks = NULL, labels = NULL, name = NULL) +
	coord_cartesian(xlim = c(0, max(x)))


```


\

Again, to help us understand the TQIP boxplots, let's flip this plot on its side.

```{r histodot-on-side}
#histo dot turned on its side
ggplot(mdata, mapping = aes(x=x)) +
	geom_dotplot(
		fill = alpha("gray", 0.5),
		color = alpha("gray", 0.5),
		binwidth = 0.1, 
		dotsize = 0.25,
		method = "histodot"
	) +
	labs(x = "Odds Ratio") +
	scale_y_continuous(breaks = NULL, labels = NULL, name = NULL) +
	scale_x_continuous(limits = c(0, max(x))) +
	coord_flip(clip="off") 

```

\

And let's rearrange all those points so they are a little easier to see. In this next plot, the points are going to be centered along an invisible vertical line and mirrored on either edge of the invisible line. 

Notice that the data haven't changed at all. We're only changing how the data are plotted.


```{r histodot-to-beeswarm}
#histodot -> bee swarm

lab = "The thick part here corresponds to taller bars on the histogram (thicker = more common)"

lab2 = "There are few centers with\nodds ratios near 0.4, so the\nplot is thin here\n(thinner = less common)"

ggplot(mdata, mapping = aes(x=1, y=x)) +
	geom_quasirandom(color = alpha("gray", 0.5)) +
	labs(y = "Odds Ratio", x = NULL) +
	scale_y_continuous(limits = c(0, max(x))) +
	scale_x_continuous(
		limits = c(0, 2),
		breaks = NULL, labels = NULL, name = NULL) +
	

	geom_curve(
		aes(x = 1.7, y = 1.5, xend = 1.5, yend = 1.05),
		arrow = arrow(
			length = unit(0.03, "npc"), 
			type="closed" # Describes arrow head (open or closed)
		),
		color = "#6EE2FF",
		size = 1.2,
		angle = 90, # Anything other than 90 or 0 can look unusual
		curvature = -0.5
	) +
		annotate(geom = "label", label = str_wrap(lab, 25),
					 x = 1.7, y = 1.6, size = 3.5,
					 fill = "#6EE2FF"
	) +
	

		geom_curve(
		aes(x = 0.35, y = 0.3, xend = 0.85, yend = 0.4),
		arrow = arrow(
			length = unit(0.03, "npc"), 
			type="closed" # Describes arrow head (open or closed)
		),
		color = "#6EE2FF",
		size = 1.2,
		angle = 75,
		curvature = -0.2
	) +
		annotate(geom = "label", label = lab2,
					 x = 0.35, y = 0.15, size = 3.5,
					 fill = "#6EE2FF"
	) 





```



\




And finally, we're going to zoom out a little bit to give us some more room to work.

```{r zoom-out}

#move to the side for later annotation 
ggplot(mdata, mapping = aes(x=0.5, y=x)) +
	geom_quasirandom(color = alpha("gray", 0.5)) +
	labs(y = "Odds Ratio", x = NULL) +
	scale_y_continuous(limits = c(0, max(x))) +
	scale_x_continuous(
		limits = c(0, 4),
		breaks = NULL, labels = NULL, name = NULL)

```


\



# The reference line

As I mentioned before, when TQIP does its benchmark reports, it provides odds ratios (AKA observed to expected ratios), and an odds ratio of one means there were just as many events as there were predicted events. 

Because of the importance of odds ratios in the benchmark report, TQIP includes a horizontal `reference line` on its plots to mark an odds ratio of one.  TQIP's plots also contain dotted lines at odds ratios of 0.5 and 2.0, but I'm not going to show those here. 


```{r ref-line}

#add reference line (OR == 1)
p = ggplot(mapping = aes(x=0.5, y=x)) +
	annotate("segment", x = -0.1, xend = 1.2, y = 1, yend = 1,
					 color="black", size = 0.75) + 
	annotate("text", x = 1.75, y = 1, label = "Ref. line", 
					 hjust = 0.5, size = 4) +
	geom_quasirandom(color = alpha("gray", 0.5)) +
	labs(y = "Odds Ratio", x = NULL) +
	scale_y_continuous(limits = c(0, max(x))) +
	scale_x_continuous(breaks = NULL, labels = NULL, name = NULL) +
	coord_cartesian(xlim = c(0, 4))



p +
	annotate(geom = "label", label = "More events than expected\nbased on the data\n(bad)",
				 x = 2.1, y = 1.35, size = 3.5,
					 fill = "#6EE2FF"
	) +
	geom_segment(
		aes(x = 1.1, xend = 1.1, y = 1.02, yend = 1.7),
		arrow = arrow(
			length = unit(0.03, "npc"), 
			type="closed" # Describes arrow head (open or closed)
		),
		color = "#6EE2FF",
		size = 1.25
	) + 
		annotate(geom = "label", label = "Fewer events than expected\nbased on the data\n(good)",
					 x = 2.1, y = 0.65, size = 3.5,
					 fill = "#6EE2FF"
	) +
	geom_segment(
		aes(x = 1.1, xend = 1.1, y = 0.98, yend = 0.3),
		arrow = arrow(
			length = unit(0.03, "npc"), 
			type="closed" # Describes arrow head (open or closed)
		),
		color = "#6EE2FF",
		size = 1.25
	)
	


```


\





# Median

In addition to giving information about the absolute value of your center's odds ratio, TQIP provides a great deal of information about *other centers'* odds ratios.  The first piece of information TQIP provides about other centers is the median odds ratio.  

The `median odds ratio` is the value at which half of all centers have odds ratios that are less than or equal to it and half of all centers have odds ratios greater than it.  The median is also known as the 50^th^ percentile since 50% of the data fall below the median and 50% falls above it.  

As you can see in this plot, the median is just a little bit greater than one (`r scales::comma(median(x), 0.01)` to be precise). So, 50% of centers had an odds ratio less than or equal to `r scales::comma(median(x), 0.01)`.


```{r median}

#add median
med = median(x)

p = p +
	annotate("segment", x = 0, xend = 1, 
					 y = med, yend = med,
					 color="gray40", size = 1.5) + 
	annotate("text", x = 1.75, y = med, 
					 label = "Median", 
					 color="gray40", size = 4, 
					 hjust = 0.5, vjust = 0.1)



#devtools::install_github("nicolash2/ggbrace")
library(ggbrace)

p + 
	annotate(geom = "label", label = "50% of centers are\nabove the median",
					 x = 2, y = 1.44, size = 3.5,
					 fill = "#6EE2FF"
	) +
	geom_brace(xstart=1, xend=1.3, 
						 ystart=med, yend=max(x),
						 color = "#6EE2FF",
						 size = 0.8,
						 rotate = 90
	) + 
	annotate(geom = "label", label = "50% of centers are\nat or below the median",
					 x = 2, y = 0.53, size = 3.5,
					 fill = "#6EE2FF"
	) +
	geom_brace(xstart=1, xend=1.3, 
						 ystart=med, yend=min(x),
						 color = "#6EE2FF",
						 size = 0.8,
						 rotate = 90
	)





```


\


# Quartiles 

In addition to the median (or 50^th^ percentile), TQIP also tells you about the 25^th^ and the 75^th^ percentiles. Just like the 50^th^ percentile, the 25^th^ percentile is simply the value at which 25% of trauma centers have odds ratios less than or equal to it.  As a result, 75% of trauma centers have odds ratios greater than the 25^th^ percentile.

As you may have guessed, the 75^th^ percentile is the point at which 75% of trauma centers have odds ratios less than or equal to it, but only 25% of trauma centers have odds ratios greater than it. 

Because a *quarter* of the data is less than or equal to the 25^th^ percentile and a *quarter* of the data is greater than the 75^th^ percentile, these values are also known as `quartiles`. Specifically, the 25^th^ percentile is the first quartile, and the 75^th^ percentile is the third quartile.  What about the second quartile though?  That's the median!  

The breakdown is that 25% of the data are at or below the first quartile; 50% of the data are at or below the second quartile; 75% of the data are at or below the third quartile; and the fourth quartile is just the maximum value so that 100% of the data is less than or equal to it. 



```{r quartiles}

#add 25th and 75th percentiles
q1 = quantile(x, probs = 0.25)
q3 = quantile(x, probs = 0.75)

p = p +
	annotate("segment", x = 0, xend = 1, 
					 y = q3, yend = q3,
					 color="gray40", size = 1.5) + 
	annotate("text", x = 1.75, y = q3, 
					 label = "75^th~percentile", parse=T,
					 color="gray40", size = 4, 
					 hjust = 0.5) +
	
	annotate("segment", x = 0, xend = 1, 
					 y = q1, yend = q1,
					 color="gray40", size = 1.5) + 
	annotate("text", x = 1.75, y = q1, 
					 label = "25^th~percentile", parse=T,
					 color="gray40", size = 4, 
					 hjust = 0.5)
	


p + 
	annotate(geom = "label", label = "25% of centers are\nabove the 3rd quartile",
					 x = 2, y = mean(c(q3, max(x))), size = 3.5,
					 fill = "#6EE2FF"
	) +
	geom_brace(xstart=1, xend=1.3, 
						 ystart=q3, yend=max(x),
						 color = "#6EE2FF",
						 size = 0.8,
						 rotate = 90
	) + 
	annotate(geom = "label", label = "75% of centers are\nat or below the 3rd quartile",
					 x = 2.1, y = mean(c(q3, min(x))), size = 3.5,
					 fill = "#6EE2FF"
	) +
	geom_brace(xstart=1, xend=1.3, 
						 ystart=q3, yend=min(x),
						 color = "#6EE2FF",
						 size = 0.8,
						 rotate = 90
	) + 
	labs(caption = "1st quartile = 25th percentile; 2nd quartile = 50th percentile = median;\n3rd quartile = 75th percentile; 4th quartile = 100th percentile = maximum")




```

\



# 10th and 90th percentiles

Briefly, TQIP also shows the 10^th^ percentile (10% of centers have odds ratios less than or equal to this point) and the 90^th^ percentile (90% of centers have odds ratios less than or equal to this point; 10% have odds ratios that are higher). 



```{r p10-and-p90}

#add 10th and 90th percentiles
p90 = quantile(x, probs = 0.9)
p10 = quantile(x, probs = 0.1)

p = p +
	annotate("segment", x = 0.2, xend = 0.8, 
					 y = p90, yend = p90,
					 color="gray40", size = 1.5) + 
	annotate("text", x = 1.75, y = p90, 
					 label = "90^th~percentile", parse=T,
					 color="gray40", size = 4, 
					 hjust = 0.5) +
	
	annotate("segment", x = 0.2, xend = 0.8, 
					 y = p10, yend = p10,
					 color="gray40", size = 1.5) + 
	annotate("text", x = 1.75, y = p10, 
					 label = "10^th~percentile", parse=T,
					 color="gray40", size = 4, 
					 hjust = 0.5)

p

```

\



# Top and bottom deciles

Just as quartiles divide the data into quarters, `deciles` divide the data into tenths. So, trauma centers that are in the first decile (the bottom decile) have odds ratios that are *lower* than 90% of TQIP trauma centers. That is, their odds ratios are less than the 10^th^ percentile. These centers are shown with green dots in the plot.  Even though this is called the bottom decile, this is where the top performing trauma centers are.  Calling it the 'bottom decile' only refers to the location on the page.  Low percentiles and low deciles are good things; that's why they are shown in green. 

Inversely, trauma centers that are in the tenth decile (the top decile) have odds ratios that are *higher* than 90% of TQIP trauma centers. That is, their odds ratios are greater than the 90^th^ percentile. These centers are shown with red dots in the plot. This is where the lowest performing 10% of trauma centers are.  High percentiles and high deciles are bad things; that's why they are shown in red.  


```{r deciles}
#add top and bottom deciles

library(gginnards) #to edit/reorder geom layers

p2 = delete_layers(p, match_type = "GeomPoint")
p2 = append_layers(p2, 
									 geom_quasirandom(
									 	data=mdata, 
									 	mapping = aes(x=0.5, y=x),				 
									 	color = mdata$color),
									 position = 2L
									 )

p2 = p2 + 
	#10th decile
	annotate("text", x = 1.75, y = max(x)*0.9, 
					 label = "Top~(10^th)~decile", parse=T,
					 color=alpha("red", 0.7), size = 4, 
					 hjust = 0.5) +
	
	#1st decile
	annotate("text", x = 1.75, y = max(x)*0.1, 
					 label = "Bottom~(1^st)~decile", parse=T,
					 color=alpha("#35B779FF", 1), size = 4, 
					 hjust = 0.5) 


p2

```




\


# Putting it all together

Now that we've covered nearly all of the elements in a TQIP boxplot, lets bring it all together to see how the data/plot we've been looking at so far translates to TQIP's boxplots. 

```{r putting-it-together}
#make TQIP boxplot

p3 = delete_layers(p2, idx=2) #delete the "Ref. line" label  
p3 = delete_layers(p3, idx=1) #delete the reference line

#add a new reference line that extends all the way across
p3 = append_layers(p3, 
									 annotate("segment", x = -0.2, xend = 4, y = 1, yend = 1,
									 				 color="black", size = 0.5),
									 position = "bottom"
)






p3 = p3 + 
	
	#50th-75th percentile
	annotate("rect", xmin = 2.5, xmax = 3.5, 
					 ymin = med, ymax = q3,
					 fill="#dedded", color = "gray40") + 
	
	#25th-50th percentile
	annotate("rect", xmin = 2.5, xmax = 3.5, 
					 ymin = q1, ymax = med,
					 fill="#dedded", color = "gray40") +
	
	#75th-90th percentile
	annotate("rect", xmin = 2.6, xmax = 3.4, 
					 ymin = q3, ymax = p90,
					 fill="#e9e9f2", color = "gray40") +	

	#10th-50th percentile
	annotate("rect", xmin = 2.6, xmax = 3.4, 
					 ymin = q1, ymax = p10,
					 fill="#e9e9f2", color = "gray40") +	
	
	#10th (top) decile
	annotate("rect", xmin = 2.7, xmax = 3.3, 
					 ymin = p90, ymax = max(x),
					 fill = "#ffdce0", color = "#ffdce0") +	
	
	#1st (bottom) decile
	annotate("rect", xmin = 2.7, xmax = 3.3, 
					 ymin = p10, ymax = min(x),
					 fill = "#a8e7b2", color = "#a8e7b2") +
	
	#75th percentile line
	annotate("segment", x = 2.5, xend = 3.5, 
					 y = q3, yend = q3,
					 color="gray40", size = 1.5) + 
	
	#25th percentile line
	annotate("segment", x = 2.5, xend = 3.5, 
					 y = q1, yend = q1,
					 color="gray40", size = 1.5) + 
	
	#median line
	annotate("segment", x = 2.5, xend = 3.5, 
					 y = med, yend = med,
					 color="gray40", size = 1.5) + 	
	
	#90th percentile line
	annotate("segment", x = 2.6, xend = 3.4, 
					 y = p90, yend = p90,
					 color="gray40", size = 1.5) + 
	
	#10th percentile line
	annotate("segment", x = 2.6, xend = 3.4,
					 y = p10, yend = p10,
					 color="gray40", size = 1.5)

	
	
	
	
p3


```



\



# Confidence intervals

Just about the only part of TQIP's plots we haven't covered yet is your center's specific odds ratio and the confidence interval that goes with it.  

Since odds ratios were discussed above, I'm going to focus on the confidence intervals here.  There is *a lot* of technical stuff that I could go into here (and maybe I will at another point), but I'm going to stick to what the confidence intervals mean for the TQIP benchmark reports. 

While the odds ratios that TQIP provides in its benchmark reports are the single best estimate for how well your center is doing, that number is still only an estimate based on TQIP's models.  The "true" odds ratio for your center is something that can never really be known with absolute certainty.  *But* with the help of some fancy math, TQIP can provide a range of likely values for your "true" odds ratio. This is where the confidence interval comes in.  

The `confidence interval` tells you that---even though no one can know the "true" odds ratio for your center---TQIP is confident that it falls within the bounds of the confidence interval. (The exact amount of confidence TQIP has will vary between reports and cohorts, so review the reference materials that come with your benchmark report to see what degree of confidence TQIP is using for each risk-adjusted metric.)

The confidence intervals are shown by vertical lines that stick out from the diamond-shaped odds ratio and then flatten out.  The points where they flatten out into horizontal lines are the upper and lower limits of the confidence interval. 


```{r}

pt = 0.92
p3 +
	geom_point(aes(x = 0.5, y = pt), shape = 18, size = 7) + 
	geom_errorbar(aes(x = 0.5, ymin = pt-0.1, ymax = pt+0.1), 
								width = 0.1) + 
	geom_point(aes(x = 3, y = pt), shape = 18, size = 7) + 
	geom_errorbar(aes(x = 3, ymin = pt-0.1, ymax = pt+0.1), 
								width = 0.1) +

	geom_curve(
		aes(x = 4, y = 1.5, xend = 3.11, yend = pt+0.1),
		arrow = arrow(
			length = unit(0.03, "npc"), 
			type="closed" # Describes arrow head (open or closed)
		),
		colour = "#6EE2FF",
		size = 1,
		angle = 90,
		curvature = -0.5
	) + 
	annotate(geom = "label", 
					 label = "Upper limit of the\nconfidence interval",
					 x = 4, y = 1.5, hjust = 1, vjust = 0,
					 fill = "#6EE2FF"
					 )

	
	

```



## Average performers: scenario #1

The importance of confidence intervals to your benchmark reports is that as long as your confidence interval crosses the reference line in the plot (the confidence interval contains 1), TQIP can be reasonably confident that your center isn't doing a lot better or a lot worse than would be expected by the data. 

In that case, your center will be an average performer, which might look something like this. Here, we can see that this center is doing a little better than expected (odds ratio of 0.92; fourth decile).  But, because the confidence interval crosses the reference line, TQIP wouldn't classify the center as a high outlier or a low outlier, just an average performer. Therefore, the diamond for the odds ratio and the bars for the confidence interval are shown in black. 


```{r ci-avg}
#confidence intervals


find_decile = function(pt){
	
	qs = quantile(x, seq(0.1,1, 0.1))
	
	res = case_when(
		pt <= qs[1] ~ 1,
		pt <= qs[2] ~ 2,
		pt <= qs[3] ~ 3,
		pt <= qs[4] ~ 4,
		pt <= qs[5] ~ 5,
		pt <= qs[6] ~ 6,
		pt <= qs[7] ~ 7,
		pt <= qs[8] ~ 8,
		pt <= qs[9] ~ 9,
		pt <= qs[10] ~ 10
	)
	
	res
	
}




#average
pt = 0.92
p3 +
	geom_point(aes(x = 0.5, y = pt), shape = 18, size = 7) + 
	geom_errorbar(aes(x = 0.5, ymin = pt-0.1, ymax = pt+0.1), 
								width = 0.1) + 
	geom_point(aes(x = 3, y = pt), shape = 18, size = 7) + 
	geom_errorbar(aes(x = 3, ymin = pt-0.1, ymax = pt+0.1), 
								width = 0.1) +
	annotate("text", x = 3, y = 0.02, vjust = 2,
					 label = paste("OR =", scales::comma(pt, 0.01))
					 ) +
	annotate("text", x = 3, y = 1.68, vjust = -4,
					 label = paste("Decile =", find_decile(pt))
	) + 
	
	geom_curve(
		aes(x = 1.7, y = 0.5, xend = 0.56, yend = 1.01),
		arrow = arrow(length = unit(0.03, "npc"), type="closed"),
		colour = "#6EE2FF", size = 1,
		angle = 90, curvature = 0.4
	) + 
	geom_curve(
		aes(x = 1.7, y = 0.5, xend = 2.96, yend = 1.01),
		arrow = arrow(length = unit(0.03, "npc"), type="closed"),
		colour = "#6EE2FF", size = 1,
		angle = 90, curvature = -0.4
	) + 
	annotate(geom = "label", 
					 label = "The upper limit of the\nconfidence interval\nis just above 1",
					 x = 1.8, y = 0.5, hjust = 0.5, vjust = 0,
					 fill = "#6EE2FF"
	)


```

\


## Average performers: scenario #2


Your odds ratio will always fall within your confidence interval. (If it doesn't, contact TQIP because something has gone very wrong!)  But the size of the confidence interval (how wide it is) will depend on a few factors.  The main factor is simply how many of your patients were in the cohort in the plot. 

This means that even though your center might be doing really well on some outcome according to the odds ratio and decile, if you don't have many patients in the cohort, TQIP might not label your center as a low outlier.  This is simply because the fewer patients you have in your data, the more uncertain TQIP's estimates will be.  And if your confidence interval crosses the reference line, your center is statistically indistinguishable from an average TQIP center. 



```{r ci-avg2, fig.height=5.5}
#average 2
pt = 0.6
p3 +
	geom_point(aes(x = 0.5, y = pt), shape = 18, size = 7) + 
	geom_errorbar(aes(x = 0.5, ymin = pt-0.5, ymax = pt+0.5), 
								width = 0.1) + 
	geom_point(aes(x = 3, y = pt), shape = 18, size = 7) + 
	geom_errorbar(aes(x = 3, ymin = pt-0.5, ymax = pt+0.5), 
								width = 0.1) +
	annotate("text", x = 3, y = 0.02, vjust = 2,
					 label = paste("OR =", scales::comma(pt, 0.01))
	) +
	annotate("text", x = 3, y = 1.68, vjust = -4,
					 label = paste("Decile =", find_decile(pt))
	) + 
	labs(caption = str_wrap("This center didn't submit many patients to TQIP. Even though their odds ratio and decile are very good, they aren't a low outlier because their cohort was small, which made TQIP's estimates less precise, and that made the confidence interval very wide."))

```


\

## High outliers (low performers)

If the lower bound of your center's confidence interval is greater than one, TQIP will label your center as a `high outlier`. This means that even after accounting for the uncertainty in the data, your center was performing worse than expected based your data. This could mean you have a quality of care issue, but it could also just mean that you have a quality of data issue. 

```{r ci-high-outlier, fig.height=5.5}
#high outlier
pt = 1.2
p3 +
	geom_point(aes(x = 0.5, y = pt), shape = 18, size = 7,
						 color = "firebrick") + 
	geom_errorbar(aes(x = 0.5, ymin = pt-0.1, ymax = pt+0.1), 
								width = 0.1, 
								color = "firebrick") + 
	geom_point(aes(x = 3, y = pt), shape = 18, size = 7, 
						 color = "firebrick") + 
	geom_errorbar(aes(x = 3, ymin = pt-0.1, ymax = pt+0.1), 
								width = 0.1, 
								color = "firebrick") +
	annotate("text", x = 3, y = 0.01, vjust = 2,
					 label = paste("OR =", scales::comma(pt, 0.01))
	) +
	annotate("text", x = 3, y = 1.7, vjust = -4,
					 label = paste("Decile =", find_decile(pt))
	) +
	labs(caption = str_wrap("This center is a high outlier/low performer because the lower bound of its confidence interval is greater than one.")) + 
		geom_curve(
		aes(x = 1.7, y = 0.5, xend = 0.58, yend = pt-0.1),
		arrow = arrow(length = unit(0.03, "npc"), type="closed"),
		colour = "#6EE2FF", size = 1,
		angle = 90, curvature = 0.4
	) + 
	geom_curve(
		aes(x = 1.7, y = 0.5, xend = 2.92, yend = pt-0.1),
		arrow = arrow(length = unit(0.03, "npc"), type="closed"),
		colour = "#6EE2FF", size = 1,
		angle = 90, curvature = -0.4
	) + 
	annotate(geom = "label", 
					 label = "The lower limit of the\nconfidence interval\nis above 1",
					 x = 1.8, y = 0.5, hjust = 0.5, vjust = 0,
					 fill = "#6EE2FF"
	)


```


\

## Low outliers (high performers)

Just as with high outliers, TQIP will let you know if you are a low outlier/high performer.  

If the upper bound of your center's confidence interval is less than one, TQIP will label your center as a `low outlier`. This means that even after accounting for the uncertainty in the data, your center was performing better than expected based your data. This could mean you provide exemplary quality of care...but it could also just mean that you have a data quality issue. 


```{r ci-low-outlier, fig.height=5.5}
#low outlier
pt = 0.85
p3 +
	geom_point(aes(x = 0.5, y = pt), shape = 18, size = 7,
						 color = "green4") + 
	geom_errorbar(aes(x = 0.5, ymin = pt-0.1, ymax = pt+0.1), 
								width = 0.1, 
								color = "green4") + 
	geom_point(aes(x = 3, y = pt), shape = 18, size = 7, 
						 color = "green4") + 
	geom_errorbar(aes(x = 3, ymin = pt-0.1, ymax = pt+0.1), 
								width = 0.1, 
								color = "green4") +
	annotate("text", x = 3, y = 0.01, vjust = 2,
					 label = paste("OR =", scales::comma(pt, 0.01))
	) +
	annotate("text", x = 3, y = 1.7, vjust = -4,
					 label = paste("Decile =", find_decile(pt))
	) +
	labs(caption = str_wrap("This center is a low outlier/high performer because the upper bound of its confidence interval is less than one."))


```



# Why some plots have different shapes  

Lastly, I'm going to discuss why it is that some cohorts have boxplots that are shaped differently from other cohorts.  For example, in many of TQIP's plots, you'll likely notice that the boxplots for some cohorts are more scrunched up than other cohorts.  


```{r}

#make a function to create my TQIP boxplots
suppressPackageStartupMessages(library(rlang))

options(dplyr.summarise.inform = FALSE)
tqip_plot = function(data, group, y, coef = 0.9, print_data=TRUE){

	
	sum_data = data %>% 
		group_by({{group}}) %>% 
		dplyr::summarise(
			min = min({{y}}, na.rm = T),
			p10 = quantile({{y}}, probs = 0.1, na.rm = T) %>% unname(),
			q1 = quantile({{y}}, probs = 0.25, na.rm = T) %>% unname(),
			med = median({{y}}, na.rm = T),
			q3 = quantile({{y}}, probs = 0.75, na.rm = T) %>% unname(),
			p90 = quantile({{y}}, probs = 0.9, na.rm = T) %>% unname(),
			max = max({{y}}, na.rm = T)
		) %>% 
		ungroup() 
	
	if(print_data){print(sum_data)}
	
	
	sum_data = sum_data %>% mutate(my_x_var = 1:n())
	
	if(is.null(coef)){coef = 0.9}

	
	ggplot(sum_data) +
		geom_hline(yintercept = 1, color="black", size = 0.75) + 
		scale_y_continuous(limits = c(min(c(floor(sum_data$min), 0)),
																	ceiling( max(sum_data$max) ))) +
		scale_x_continuous(
			breaks = sum_data$my_x_var,
			labels = sum_data %>% pull({{group}})
		) +
		#50th-75th percentile
		geom_rect(aes(xmin = my_x_var-(0.5*coef) , xmax = my_x_var+(0.5*coef),
									ymin = med, ymax = q3),
							fill="#dedded", color = "gray40"
		) + 
		
		#25th-50th percentile
		geom_rect(aes(xmin = my_x_var-(0.5*coef), xmax = my_x_var+(0.5*coef),
									ymin = q1, ymax = med),
							fill="#dedded", color = "gray40"
		)   + 
		
		
		#75th-90th percentile
		geom_rect(aes(xmin = my_x_var-(0.4*coef), xmax = my_x_var+(0.4*coef),
									ymin = q3, ymax = p90),
							fill="#e9e9f2", color = "gray40"
		)   + 
		
		
		#10th-50th percentile
		geom_rect(aes(xmin = my_x_var-(0.4*coef), xmax = my_x_var+(0.4*coef),
									ymin = q1, ymax = p10),
							fill="#e9e9f2", color = "gray40"
		)   + 
		
		
		#10th (top) decile
		geom_rect(aes(xmin = my_x_var-(0.3*coef), xmax = my_x_var+(0.3*coef),
									ymin = p90, ymax = max),
							fill="#ffdce0", color = "#ffdce0"
		)   + 
		
		#1st (bottom) decile
		geom_rect(aes(xmin = my_x_var-(0.3*coef), xmax = my_x_var+(0.3*coef),
									ymin = p10, ymax = min),
							fill="#a8e7b2", color = "#a8e7b2"
		)   + 
		
		
		
		#75th percentile line
		geom_segment(aes(x = my_x_var-(0.5*coef), xend = my_x_var+(0.5*coef),
										 y = q3, yend = q3),
								 color="gray40", size = 1.5*coef) +
		
		
		#25th percentile line
		geom_segment(aes(x = my_x_var-(0.5*coef), xend = my_x_var+(0.5*coef),
										 y = q1, yend = q1),
								 color="gray40", size = 1.5*coef)  +
		
		#median line
		geom_segment(aes(x = my_x_var-(0.5*coef), xend = my_x_var+(0.5*coef),
										 y = med, yend = med),
								 color="gray40", size = 1.5*coef) +
		
		
		#90th percentile line
		geom_segment(aes(x = my_x_var-(0.4*coef), xend = my_x_var+(0.4*coef),
										 y = p90, yend = p90),
								 color="gray40", size = 1.5*coef) +
		
		#10th percentile line
		geom_segment(aes(x = my_x_var-(0.4*coef), xend = my_x_var+(0.4*coef),
										 y = p10, yend = p10),
								 color="gray40", size = 1.5*coef) +
		labs(
			x = enquo(group),
			y = enquo(y)
		)
	
	
	
}



set.seed(48)
temp = tibble(
	or = c(rnorm(100, 1, sd = 0.3), rnorm(100, 1, sd = 0.1), rnorm(100, 1, sd = 0.5)),
	cohort = rep(c("A", "B", "C"), each = 100)
)


set.seed(1)
temp = bind_rows(temp,
								 tibble(
								 	or = rnorm(100, 1.2, sd = 0.3),
								 	cohort = rep("D", 100)
								 )
)

# tapply(temp$or, temp$cohort, summary)

temp %>% tqip_plot(group = cohort, y = or, print_data=F) + 
	labs(y = "Odds Ratio", x = "Cohort")



```


This can happen when the distribution of odds ratios differs between cohorts.  In the above example, all the TQIP centers had odds ratios very close to 1 for Cohort B. Since all the centers were doing about as well as expected, the boxplot for that cohort is very tightly centered around the reference line (1.00).  However, there was a lot more variability in the performance of centers when it came to Cohort C. As a result, the boxplot is much more spread out in that cohort. 


Adding the dots for centers' individual odds ratios back in can help make this more clear. 


```{r}


pt = temp %>% 
	tqip_plot(group = cohort, y = or, print_data=F) +
	labs(y = "Odds Ratio", x = "Cohort")




temp = temp %>% mutate(
		x = cohort %>% as.factor() %>% as.numeric()
	) 
	
pt + 	
	geom_quasirandom(
		data = temp, 
		mapping = aes(x = x, y = or),
		color = alpha("black", 1)
		)


```




# Conclusion  

Hopefully this inaugural post on the Trauma Data Blog helped clear up a lot of questions you had about TQIP's boxplots in the benchmark reports. If it didn't or you still have questions, send me an [email](mailto:TraumaDataBlog@gmail.com) or leave a comment.  I'd also love suggestions for future topics. 


